---
title: "iNaturalist data in R"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---


Goals:

* Query iNaturalist data directly from RStudio, using different parameters
* Calculate Diversity indices (Shannon, Simpson, Species Richness) from iNaturalist data
* Show a couple data visualizations (barplot & map)


In this workshop, we're going to use the "rinat" library (probably pronounced "R-iNat"). You can query data directly from iNaturalist with rinat, and then use R scripting to target data columns, run calculations, generate charts, etc.


First let's load the rinat, tidyverse, and vegan libraries:

```{r}
# for interacting with iNaturalist
library(rinat)

# working with data frames & charts
library(tidyverse)

# for calculating diversity indices
library(vegan)

```


The most useful function in rinat is get_inat_obs (more here: https://rdrr.io/cran/rinat/man/get_inat_obs.html). 

Let's do a quick search:

```{r}

spiders<-get_inat_obs(query="Arachnids")

spiders

```

This is great....we now have a data frame of 100 observations of Arachnids. However, the results are a little scattered geographically, and we only have 100, so may be of limited use. If we want to do some real analysis of data, we should ask a more targeted question using the available get_inat_obs arguments:


query:	Query string for a general search.

taxon_name: Filter by iNat taxon name. Note that this will also select observations of descendant taxa. Note that names are not unique, so if the name matches multiple taxa, no observations may be returned.

taxon_id: Filter by iNat taxon ID. Note that this will also select observations of descendant taxa.

place_id: Filter by iNat place ID.

quality: The quality grade to be used. Must be either "casual" or "research". If left blank both will be returned.

geo: Flag for returning only results that are georeferenced, TRUE will exclude non-georeferenced results, but they cannot be excluded.

year: Return observations only in that year (can only be one year, not a range of years).

month: Return observations only by month, must be numeric, 1...12

day: Return observations only on a given day of the month, 1...31

bounds: A bounding box of longitude (-180 to 180) and latitude (-90 to 90) to search within. It is a vector in the form of southern latitude, western longitude, northern latitude, and eastern longitude. Alternatively supply an sf or sp object from which the bounding box will be derived.

maxresults: The maximum number of results to return. Should not be a number higher than 10000.

meta: (logical) If TRUE, the output of this function is a list with metadata on the output and a data.frame of the data. If FALSE (default), just the data.frame.

--

In order to get the most precise data, it's probably a good idea to use some of the more specific arguments (e.g. taxon_name, taxon_id, place_id), rather than the more general argument "query". Of course, that's only useful if you know what taxon_id and place_id are. Fortunately, you can find them pretty easily by using their web interface for places:

1) Visit https://www.inaturalist.org/places
2) Perform a search for a place (this can be a city, state, county, country, point of interest, etc.).
3) View the results, and click on the link for the place you want.
4) The resulting URL should give you the place id. For example, for the state of Oregon, the URL is: https://www.inaturalist.org/places/oregon-us In this case, the place id is "oregon-us"

From this website, you can also get the taxon_id. Let's say we're interested in all Arachnids. On the left side, if I click on Arachnids, I get this URL:

https://www.inaturalist.org/places/oregon-us#taxon=47119

...so my taxon_id is 47119.

You can also get the taxon_id from the URL of a taxon page in iNaturalist. For example:

https://www.inaturalist.org/taxa/67708-Argiope-trifasciata

In the case above, the taxon_id is 67708.

Let's say I want to get all the research grade results of Arachnids in Oregon for the year 2019. I can enter my arguments in get_inat_obs below (keep in mind, it may take several seconds to get the results):

```{r}

#make the query
#the default maxresults is 100, so bumping it up to 10000 (the max) should give you a larger result set

orArachnids2019<-get_inat_obs(place_id="oregon-us", taxon_id="47119", quality="research", year="2019", maxresults = 10000)

#see the results
orArachnids2019
```

This is a much more useful dataset. Below we'll use this to calculate diversity indices, and create some maps & charts.

But first...
Can you get a data set from iNaturalist of all research level Arachnids observations in the Portland-Vancouver metro region from 2019? (Hint: start at https://www.inaturalist.org/places to get the place_id and taxon_id)

```{r}



```


Uploading data
Despite the awesomeness of rinat, you might choose to download your data directly from iNaturalist. Doing this extends your dataset beyond 10,000, and you can also get additional columns that could be useul to your project ("taxon extras", like "taxon_kingdom_name", etc.). You can still use this data in RStudio, by doing the following:

* Download and unzip your dataset from iNaturalist. This should give you a .csv file.
* In RStudio, click "Upload" in the Files area (lower right-hand corner). Select your file, and click Ok.
* Once your file has been uploaded, use the read_csv function to create a data frame variable:

myData<-read_csv("filename.csv")
myData

If you have a dataset, try this below:

```{r}



```


### Diversity Indices


Now let's revisit the Arachnid dataset to determine the Shannon, Simpson, and Species Richness diversity indices. 


First let's calculate Species Richness (count of all species). For this, we can use the "unique" function, which looks for all unique values in a data frame column. Then we use the "length" function to count how many values were calculated by "unique". This will give us the Species Richness:

```{r}
# calculate species richness:
# Find unique values in the scientific_name column
vals<-unique(orArachnids2019$scientific_name)
#print it
vals

# length gives to total number (Species Richness)
length(vals)

```



For the Shannon and Simpson indices, we need to run some code to group our data by species, and determine the respective species counts (how many unique species are represented, and how many of each).  To accomplish this, we're going to use the "pipe" operator: %>%. This sort of means "now pass this on to the next function. 

"group_by" will group the data by unique values in the column I specify. Then once it's grouped, the "summarize" function lets me create columns based upon some criteria or calculation. In the example below, I'm grouping by scientific_name first, and then using summarize to create a column "n", equal to the count ("n()") of each group.



```{r}

speciesCounts<-orArachnids2019 %>% group_by(scientific_name) %>% summarize(n=n())

speciesCounts


```



Now lets say we want to calculate the Shannon index for species diversity. Fortunately, the "vegan" package has a function called "diversity" that will do this in a snap. We just specify the column of numeric data (n)


```{r}
shannonIndexArachnids<-diversity(speciesCounts$n, index="shannon")

shannonIndexArachnids

```

Similarly, we can calculate the Simpson index:

```{r}

simpsonIndexArachnids<-diversity(speciesCounts$n, index="simpson")

simpsonIndexArachnids


```

Now, calculate Simpson and Shannon indicies for the Portland-Vancouver Metro Region Arachnids dataset:


```{r}

# Species Richness:
# use "unique" and "length"


# Shannon & Simpson:
# first create a dataset of species & their counts, using group_by & summarize:


# now use the diversity function to generate and print the indices:



```







## Visualizations:

We're just going to scratch the surface with what kinds of visualizations are possible. As your group is deciding what kind of visualization(s) to create, do some googling, and reach out to Jeremy if you need some guidance (jeremym@lclark.edu).


### Bar charts

Let's say you want to show a bar chart comparing a Shannon indices from 2017-2020. You may have to first calculate values as described above, and then create a data frame from those values. Once that's in place, you can then plot your data. Here's an example using some completely arbitrary values:

```{r}

# Here are some pretend Shannon index values...normally you would calculate them as described above
sh2017<-4.1
sh2018<-3.8
sh2019<-6.4
sh2020<-3.3

# Create two vectors, with corresponding values in order
years <- c('2017', '2018','2019','2020')
shannonVals <- c(sh2017, sh2018, sh2019, sh2020)

# Use the data.frame function to combine the vectors
myData <- data.frame(years, shannonVals)

# Printe the the new data frame
myData


# Generate the barlot (https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/barplot)

barplot(myData$shannonVals, names.arg=myData$years, xlab="Year",ylab="Shannon Diversity Index",col="blue",
main="Shannon Indices of Arachnids in Oregon")
```



### Mapping Data

The rinat package has a function for taking datasets from iNaturalist and plotting them on a map (inat_map).

Using the Arachnids data set, let's look at its arguments:

data: a dataframe of iNaturalist results

map: Mainly indicates the political borders to display. Options are "usa", "county", "state".

subregion: Usually a state name, or "." (for "everything"), and works with the "map" value. 



```{r}

# shows all states, regardless of where data is
usaMap<-inat_map(data=orArachnids2019, map = "state", subregion = ".", plot = TRUE)

# with map set to "county", and subregrion set to "Oregon", shows just Oregon with county borders displayed
oregon<-inat_map(data=orArachnids2019, map = "county", subregion = "Oregon", plot = TRUE)
#x+labs(title="My Title")

# set plot to "false"...doesn't print, but I can add stuff to it later

oregon2<-inat_map(data=orArachnids2019, map = "county", subregion = "Oregon", plot = FALSE)
oregon2+labs(title="Arachnids in Oregon, 2019")


```














